---
title: "P8105 Homework 2"
author: "Cecilia Zhao"
output: github_document
---

```{r, message=FALSE}
library(tidyverse)
library(readxl)
```

## Problem 1

```{r, message=FALSE}
transit_df = 
  read_csv(
    "data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv",
    na = c("", "NA", ".")
  )

transit_df = janitor::clean_names(transit_df)
```

```{r}
transit_df = 
  select(
    transit_df,
    line,
    station_name,
    station_latitude,
    station_longitude,
    route1:route11,
    entry,
    vending,
    entrance_type,
    ada
  )
```

```{r, warning=FALSE}
transit_df = 
  transit_df |> 
  mutate(
    entry = case_match(
      entry,
      "YES" ~ TRUE,
      "NO" ~ FALSE
    )
  )
```

This dataset contains information about entrances and exits for NYC subway stations. It includes the subway line, station name and location, routes served, entrance type, whether entry is allowed, vending availability, and ADA compliance.
I cleaned the data by standardizing variable names, retaining only the variables relevant to the analysis, and converting the `entry` variable from a character variable (`YES` / `NO`) to a logical variable (`TRU`E / `FALSE`). After cleaning, the resulting dataset contains `r nrow(transit_df)` rows and `r ncol(transit_df)` columns.
These data are not tidy because the routes served by each station are spread across multiple columns, instead of being stored as values of a single route variable.

```{r}
n_stations = 
  transit_df |> 
  distinct(line, station_name) |> 
  nrow()
```

There are `r n_stations` distinct subway stations.

```{r}
n_ada_stations = 
  transit_df |> 
  distinct(line, station_name, ada) |> 
  filter(ada == TRUE) |> 
  nrow()
```

There are `r n_ada_stations` ADA compliant stations.

```{r}
prop_entry_no_vending = 
  transit_df |> 
  filter(vending == "NO") |> 
  summarize(
    proportion = mean(entry)
  ) |> 
  pull(proportion)
```

Among station entrances/exits without vending, `r round(prop_entry_no_vending * 100, 1)`% allow entrance.

```{r}
transit_long = 
  transit_df |> 
  pivot_longer(
    route1:route11,
    names_to = "route_number",
    values_to = "route_name",
    values_transform = as.character
  )
```

```{r}
n_a_stations = 
  transit_long |> 
  filter(route_name == "A") |> 
  distinct(line, station_name) |> 
  nrow()
```

```{r}
n_a_ada_stations = 
  transit_long |> 
  filter(route_name == "A", ada == TRUE) |> 
  distinct(line, station_name) |> 
  nrow()
```

There are `r n_a_stations` distinct stations that serve the A train. Among these stations, `r n_a_ada_stations` are ADA compliant.



## Problem 2

```{r}
mr_trash_df = 
  read_excel(
    "data/202509 Trash Wheel Collection Data.xlsx",
    sheet = "Mr. Trash Wheel",
    range = "A2:N709"
  )

mr_trash_df = janitor::clean_names(mr_trash_df)

mr_trash_df =
  mr_trash_df |>
  filter(!is.na(dumpster)) |>
  mutate(
    sports_balls = as.integer(round(sports_balls)),
    year = as.numeric(year),
    trash_wheel = "Mr. Trash Wheel"
  )
```


```{r}
prof_trash_df =
  read_excel(
    "data/202509 Trash Wheel Collection Data.xlsx",
    sheet = "Professor Trash Wheel",
    range = "A2:M134"
  )

prof_trash_df = janitor::clean_names(prof_trash_df)
```

```{r}
prof_trash_df =
  prof_trash_df |>
  filter(!is.na(dumpster)) |>
  mutate(
    trash_wheel = "Professor Trash Wheel"
  )
```

```{r}
gwynnda_df =
  read_excel(
    "data/202509 Trash Wheel Collection Data.xlsx",
    sheet = "Gwynns Falls Trash Wheel",
    range = "A2:L351"
  )

gwynnda_df = janitor::clean_names(gwynnda_df)
```

```{r}
gwynnda_df =
  gwynnda_df |>
  filter(!is.na(dumpster)) |>
  mutate(
    trash_wheel = "Gwynnda Wheel"
  )
```

```{r}
trash_df =
  bind_rows(
    mr_trash_df,
    prof_trash_df,
    gwynnda_df
  )
```

```{r}
prof_total_weight =
  trash_df |>
  filter(trash_wheel == "Professor Trash Wheel") |>
  summarize(
    total_weight = sum(weight_tons, na.rm = TRUE)
  ) |>
  pull(total_weight)
```

```{r}
gwynnda_june_2022 =
  trash_df |>
  filter(
    trash_wheel == "Gwynnda Wheel",
    year == 2022,
    month == "June"
  ) |>
  summarize(
    total_butts = sum(cigarette_butts, na.rm = TRUE)
  ) |>
  pull(total_butts)
```

The combined Trash Wheel dataset contains `r nrow(trash_df)` observations, which are come from Mr. Trash Wheel, Professor Trash Wheel, and Gwynnda Wheel.
Key variables include dumpster number, date, trash wheel, weight of trash collected, volume, plastic bottles, polystyrene, cigarette butts, plastic bags, and wrappers.
Professor Trash Wheel collected `r round(prof_total_weight, 2)` tons of trash in total.
In June 2022, Gwynnda collected a total of `r format(gwynnda_june_2022, big.mark = ",", scientific = FALSE)` cigarette butts.


## Problem 3

```{r}
baseline_df =
  read_csv(
    "data/data_mci/MCI_baseline.csv",
    skip = 1,
    na = c("", "NA", ".")
  )

baseline_df = janitor::clean_names(baseline_df)
```


```{r}
baseline_df =
  baseline_df |>
  mutate(
    sex = case_match(
      sex,
      0 ~ "Female",
      1 ~ "Male"
    ),
    apoe4 = case_match(
      apoe4,
      0 ~ "Non-carrier",
      1 ~ "Carrier"
    )
  )
```

```{r}
baseline_df =
  baseline_df |>
  filter(
    is.na(age_at_onset) | age_at_onset > current_age
  )
```

```{r}
n_participants = nrow(baseline_df)

n_mci =
  baseline_df |>
  filter(!is.na(age_at_onset)) |>
  nrow()

mean_baseline_age =
  baseline_df |>
  summarize(
    mean_age = mean(current_age, na.rm = TRUE)
  ) |>
  pull(mean_age)

prop_women_apoe4 =
  baseline_df |>
  filter(sex == "Female") |>
  summarize(
    proportion = mean(apoe4 == "Carrier")
  ) |>
  pull(proportion)
```

For the data cleaning process, I first imported the MCI_baseline csv file and skipped the first row of variable descriptions. I treatied "." as missing values. Variable names were cleaned to have a consistent format. Sex and APOE4 carrier status were recoded from numeric indicators to categorical variables based on the description on the first line. Participants who had developed MCI at or before baseline were removed to meet the study inclusion criteria. The cleaned dataset contains `r n_participants` participants. The variables in this dataset are baseline age, sex, education, APOE4 carrier status, and age at MCI onset. Among the `r n_participants` participants, `r n_mci` developed MCI during follow-up. The average baseline age was  `r round(mean_baseline_age, 2)` years, and `r round(prop_women_apoe4 * 100, 1)`% of women were APOE4 carriers.


```{r}
amyloid_df =
  read_csv(
    "data/data_mci/MCI_amyloid.csv",
    skip = 1,
    na = c("", "NA", "Na", ".")
  )

amyloid_df = janitor::clean_names(amyloid_df)
```

```{r}
amyloid_df =
  amyloid_df |>
  pivot_longer(
    baseline:time_8,
    names_to = "time",
    values_to = "amyloid"
  )
```

```{r}
amyloid_df =
  amyloid_df |>
  mutate(
    time = case_match(
      time,
      "baseline" ~ 0,
      "time_2" ~ 2,
      "time_4" ~ 4,
      "time_6" ~ 6,
      "time_8" ~ 8
    )
  )
```

```{r}
amyloid_df =
  amyloid_df |>
  rename(id = study_id)
```

```{r}
baseline_only =
  baseline_df |>
  anti_join(amyloid_df, by = "id")

amyloid_only =
  amyloid_df |>
  anti_join(baseline_df, by = "id") |>
  distinct(id)
```

```{r}
combined_df =
  amyloid_df |>
  inner_join(
    baseline_df,
    by = "id"
  )
```

```{r}
write_csv(
  combined_df,
  "data/mci_complete.csv"
)
```

First, I imported he amyloid biomarker dataset, skipped the first row of variable descriptions, treated missing-value codes as NA. The data were reshaped from wide to long format so that each row represents one participant at one measurement time. I also recoded the measurement times as 0, 2, 4, 6, and 8 years, and amyloid β 42/40 ratio values were converted to numeric. The resulting longitudinal dataset contains `r nrow(amyloid_df)` observations from `r n_distinct(amyloid_df$id)` participants.
I Compared participant IDs across the two datasets. It showes that `r n_distinct(baseline_only$id)` participants appeared only in the cleaned baseline dataset, while `r n_distinct(amyloid_only$id)` participants appeared only in the amyloid dataset. Then I used inner join to retain only participants present in both datasets. The combined dataset contains `r nrow(combined_df)` observations from `r n_distinct(combined_df$id)` and `r ncol(combined_df)` variables. These variables include participant ID, measurement time, amyloid β 42/40 ratio, baseline age, sex, education, APOE4 carrier status, and age at MCI onset. The combined dataset was exported as mci_complete.csv.

