---
title: "P8105 Homework 2"
author: "Cecilia Zhao"
output: github_document
---

```{r, message=FALSE}
library(tidyverse)
library(readxl)
```

## Problem 1

```{r, message=FALSE}
transit_df = 
  read_csv(
    "data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv",
    na = c("", "NA", ".")
  )

transit_df = janitor::clean_names(transit_df)
```

```{r}
transit_df = 
  select(
    transit_df,
    line,
    station_name,
    station_latitude,
    station_longitude,
    route1:route11,
    entry,
    vending,
    entrance_type,
    ada
  )
```

```{r, warning=FALSE}
transit_df = 
  transit_df |> 
  mutate(
    entry = case_match(
      entry,
      "YES" ~ TRUE,
      "NO" ~ FALSE
    )
  )
```

This dataset contains information about entrances and exits for NYC subway stations. It includes the subway line, station name and location, routes served, entrance type, whether entry is allowed, vending availability, and ADA compliance.
I cleaned the data by standardizing variable names, retaining only the variables relevant to the analysis, and converting the `entry` variable from a character variable (`YES` / `NO`) to a logical variable (`TRU`E / `FALSE`). After cleaning, the resulting dataset contains `r nrow(transit_df)` rows and `r ncol(transit_df)` columns.
These data are not tidy because the routes served by each station are spread across multiple columns, instead of being stored as values of a single route variable.

```{r}
n_stations = 
  transit_df |> 
  distinct(line, station_name) |> 
  nrow()
```

There are `r n_stations` distinct subway stations.

```{r}
n_ada_stations = 
  transit_df |> 
  distinct(line, station_name, ada) |> 
  filter(ada == TRUE) |> 
  nrow()
```

There are `r n_ada_stations` ADA compliant stations.

```{r}
prop_entry_no_vending = 
  transit_df |> 
  filter(vending == "NO") |> 
  summarize(
    proportion = mean(entry)
  ) |> 
  pull(proportion)
```

Among station entrances/exits without vending, `r round(prop_entry_no_vending * 100, 1)`% allow entrance.

```{r}
transit_long = 
  transit_df |> 
  pivot_longer(
    route1:route11,
    names_to = "route_number",
    values_to = "route_name",
    values_transform = as.character
  )
```

```{r}
n_a_stations = 
  transit_long |> 
  filter(route_name == "A") |> 
  distinct(line, station_name) |> 
  nrow()
```

```{r}
n_a_ada_stations = 
  transit_long |> 
  filter(route_name == "A", ada == TRUE) |> 
  distinct(line, station_name) |> 
  nrow()
```

There are `r n_a_stations` distinct stations that serve the A train. Among these stations, `r n_a_ada_stations` are ADA compliant.



## Problem 2









